CUDA-accelerated Genetic Feedforward-ANN Training for Data Mining

### 2010-04-19 doc/abstract.tex
Found another mistake
git-svn-id: http://svn.scesoc.ca/evolution/trunk@105 fee98b4c-d1c3-4d34-9525-a468159e4a48
Address JRG's comments to section Methods and following
git-svn-id: http://svn.scesoc.ca/evolution/trunk@106 fee98b4c-d1c3-4d34-9525-a468159e4a48

% ############################################################################
% Abstract
% ############################################################################
\begin{abstract}
We present an implementation of genetic algorithm (GA) training of artificial neural networks (ANNs) targeting commodity graphics cards (GPUs). By carefully mapping the problem onto the unique GPU architecture, we achieve order-of-magnitude speedup over a conventional CPU implementation. Furthermore, we show that the speedup is consistent across a wide range of data set sizes, making this implementation ideal for large data sets. This performance boost enables the genetic algorithm to search a larger subset of the solution space, which results in a more accurate pattern classifier. Finally, we demonstrate this method in the context of the 2009 UC San Diego Data Mining Contest, achieving a world-class lift on a data set of 94682 e-commerce transactions.
\end{abstract}

@@

% ----------------------------------------------------------------------------
\subsection{Artificial Neural Networks} \label{ann}
% ----------------------------------------------------------------------------
ANNs are composed of nodes connected by weighted directed edges. In a feed-forward network architecture, nodes are organized in an input layer, several hidden layers, and an output layer. Each layer is fully connected to the next. Each of the hidden nodes at hidden layer $n$ perform a transcendental function with inputs equal to the outputs of layer $n-1$ each multiplied by the weight on the edge through which they are fed. In a radial basis function (RBF) network, the nodes in the hidden layer calculate the distance between their inputs and a centre vector, and pass a weighted distance to a RBF. Nodes in a gaussian function network perform a similar operation with a gaussian function in place of the RBF. The output node is typically linear, computing a weighted sum of its inputs. The output is a single real value which is interpreted as a classification confidence. ANN computation is dominated by the number of multiplications at the edges between nodes, proportional to the number of features and the number of hidden nodes, and by the transcendental functions (exponentiation for RBF) at the hidden nodes.  ANN training has been implemented previously on GPU hardware, however these efforts have focused on training through backpropagation \cite{backprop} as opposed to genetic algorithms.

@@

% ############################################################################
\section{Methods} \label{algimp}
% ############################################################################

% ----------------------------------------------------------------------------
\subsection{Implementation of GA Training of ANNs} \label{implementation}
% ----------------------------------------------------------------------------
Each iteration of the GA (referred to as a generation) consists of the following steps:
\begin{enumerate}
	\item Computation of ANN outputs, $i.e.$ classification confidence for each instance in the data set as calculated by each candidate ANN in the generation
	\item Calculation of top 20\% threshold value for each candidate ANN
	\item Counting of the positive instances with outputs above this threshold (each individual)
	\item Genetic algorithm processing -- mating, mutation, selection
\end{enumerate}

Items 1, 2 and 3 were all implemented on the CUDA platform. Performance results are given as the combined time for one generation of these three items but exclude one-time initialization. Item 4 is not very computationally intensive and was not parallelized.

% ----------------------------------------------------------------------------
\subsection{Computation of ANN Output Values} \label{implementation}
% ----------------------------------------------------------------------------
ANN output values are computed by reading feature values and applying the appropriate operations represented by the network topology and parameters. Feature data are organized to enable memory read coalescing. This part of the process is particularly well-suited to the CUDA architecture because it has a very high ratio of mathematical operations to memory accesses (160 multiplications, 4 exponentiations and 80 bytes of memory accesses for the ANN topology used during testing). Parallelization is across individuals as well as training instances, which allows this part of the process to scale particularly well with hardware capabilities.

% ----------------------------------------------------------------------------
\subsection{Calculation of Top 20\% Threshold} \label{implementation}
% ----------------------------------------------------------------------------
To efficiently calculate the top 20\% threshold in a list of $n$ unsorted ANN output values, we can insert all values into a $k$-sized binary minheap where $k = 0.20 * n$, and take the root of the heap as the threshold. Each minheap insertion costs $O(\log k)$ time. Because device shared memory (16~KB) is not large enough for our value of $k$, we instead use $p$ passes of $\frac{k}{p}$-sized heaps with which we find the top 5\% threshold, then the top 10\%, etc. The overall complexity of our algorithm is $O(np \log \frac{k}{p})$. Parallelization is across individuals.

% ----------------------------------------------------------------------------
\subsection{Counting of Positives Above Threshold} \label{implementation}
% ----------------------------------------------------------------------------
By keeping positive instances first in our training set, we avoid the need to consult the class of each instance (global memory reads are costly). We simply iterate over the first $numPositives$ ANN outputs and count the outputs that are greater than the individual's 20\% threshold. Parallelization is across both individuals and groups of outputs. For each individual, we keep m counters which accumulate counts for output indices $i$ where $i \equiv 0 \bmod{m}$, $i \equiv 1 \bmod{m}$, ..., $i \equiv m - 1 \bmod{m}$. This enables parallelization even when there are not enough individuals in the population to fully occupy the device. Each batch of $m$ kernels accesses $m$ sequential locations in memory, which results in coalesced reads.

@@

\begin{table}[h]
	\centering
	\begin{tabular}{ll}
	\textbf{Legend Label} & \textbf{Hardware} \\
	\hline
	Core2 & Intel Core 2 Q9450 CPU \\
	i7 & Intel i7 920 CPU \\
	GTX260 & NVidia GTX260 GPU \\
	GTX275 & NVidia GTX275 GPU \\
	\hline
	\end{tabular}
	\caption{Experimental hardware}
	\label{tab:experimental-hardware}
\end{table}

% ----------------------------------------------------------------------------
\subsection{Performance Results} \label{performance}
% ----------------------------------------------------------------------------
In most cases, the GPU implementation resulted in approximately an order of magnitude speedup over a CPU implementation (Figure~\ref{fig:training-performance}). For low population sizes, GPU initialization and data copying dominates.

\begin{figure}[h]
	\includegraphics[width=\textwidth]{fig-performance}
	\caption{Training performance as a function of population size (left) and training set size (right). Note logarithmic axes.}
	\label{fig:training-performance}
\end{figure}

% ############################################################################
\section{Conclusion} \label{concl}
% ############################################################################
We have presented an implementation of GA training of ANN classifiers for the CUDA platform for GPU programming. By carefully designing memory organization, algorithm computational load and memory access patterns, we have obtained a 10-fold speedup compared to a conventional sequential CPU implementation. Additional workarounds were necessary for the low amount of fast on-chip GPU memory. Our method scales across population and training set sizes.

>>>>

% ############################################################################
% Abstract
% ############################################################################
\begin{abstract}
We present an implementation of genetic algorithm (GA) training of feedforward artificial neural networks (ANNs) targeting commodity graphics cards (GPUs). By carefully mapping the problem onto the unique GPU architecture, we achieve order-of-magnitude speedup over a conventional CPU implementation. Furthermore, we show that the speedup is consistent across a wide range of data set sizes, making this implementation ideal for large data sets. This performance boost enables the genetic algorithm to search a larger subset of the solution space, which results in a more accurate pattern classifier. Finally, we demonstrate this method in the context of the 2009 UC San Diego Data Mining Contest, achieving a world-class lift on a data set of 94682 e-commerce transactions.
\end{abstract}

@@

% ----------------------------------------------------------------------------
\subsection{Artificial Neural Networks} \label{ann}
% ----------------------------------------------------------------------------
ANNs are composed of nodes connected by weighted directed edges. In a feed-forward network architecture, nodes are organized in an input layer, several hidden layers, and an output layer. Each layer is fully connected to the next. Each of the hidden nodes at hidden layer $n$ perform a transcendental function with inputs equal to the outputs of layer $n-1$ each multiplied by the weight on the edge through which they are fed. In a radial basis function (RBF) network, the nodes in the hidden layer calculate the distance between their inputs and a centre vector, and pass a weighted distance to a RBF. Nodes in a sigmoid function network perform a similar operation with a sigmoid function in place of the RBF. The output node is typically linear, computing a weighted sum of its inputs. The output is a single real value which is interpreted as a classification confidence. ANN computation is dominated by the number of multiplications at the edges between nodes, proportional to the number of features and the number of hidden nodes, and by the transcendental functions (exponentiation for RBF) at the hidden nodes.  ANN training has been implemented previously on GPU hardware, however these efforts have focused on training through backpropagation \cite{backprop} as opposed to genetic algorithms.

@@

% ############################################################################
\section{Implementation of GA Training of ANNs} \label{algimp}
% ############################################################################

Our GA begins by generating an initial population of random individuals, each of which is a set parameters describing one ANN. Feature values for all training instances are loaded into GPU global memory in preparation for ANN computations. Then, the following steps are executed in sequence for each generation of the GA:

\begin{enumerate}
	\item Compute the ANN output corresponding to each training instance, as calculated by each candidate ANN in the population.
	\item Find the threshold that separates the highest 20\% outputs for each candidate's set of output values.
	\item Compute the number of top 20\% instances which are truly positive, again for each candidate ANN. This allows us to calculate lift, which is used as the fitness value for each candidate ANN.
	\item Apply genetic operators: mating, mutation and selection. The selected candidate ANNs become the new population, to be used in the next generation.
\end{enumerate}

Items 1, 2 and 3 were all implemented on the CUDA platform and are described in detail below. Performance results are given as the combined time for one generation of these three items but exclude one-time initialization. Item 4 is not very computationally intensive and was not parallelized.

% ----------------------------------------------------------------------------
\subsection{Computation of ANN Output Values} \label{implementation}
% ----------------------------------------------------------------------------
ANN output values are computed by reading the training instances as real-valued feature vectors and applying the appropriate operations represented by the network topology and parameters. Feature data are organized to efficiently use memory bandwidth. ANN output calculation is particularly well-suited to the CUDA architecture because it has a very high ratio of mathematical operations to memory accesses (160 multiplications, 4 exponentiations and 80 bytes of memory accesses for our ANN topology). In addition, each output value depends only on the current ANN parameters and one training instance's feature values. Therefore, parallelization of this step scales particularly well with hardware capabilities.

% ----------------------------------------------------------------------------
\subsection{Calculation of Top 20\% Threshold} \label{implementation}
% ----------------------------------------------------------------------------
Next, we must calculate the $k^{th}$ largest value in each candidate's list of output values, where $k = 0.20 * n$ and $n$ is the number of output values (number of training instances). We can accomplish this by iterating through all $n$ unsorted output values and inserting each into a binary minheap of fixed size $k$. The heap size is kept fixed by removing the root (smallest node) after each insertion. At the end of this process, the root of the heap is the $k^{th}$ largest output value. The insertion and removal can be combined into a single operation which costs $O(\log k)$ time. The heap occupies $4 * k$ bytes of device shared memory.

Because fast on-chip memory (16~KB) is not large enough for our value of $k$ ($4 * 18936 =$ 74~KB), we instead use $p$ passes of $\frac{k}{p}$-sized heaps. Each pass examines only values which are less than the threshold calculated by the preceeding pass. For example, the third pass examines only values below the 10\% threshold and finds the top 5\% threshold of this subset of values. This threshold however is the 15\% threshold of the full list. The threshold calculated by the last pass is the desired 20\% threshold.

The overall complexity of our algorithm is $O(np \log \frac{k}{p})$. Each top 20\% calculation is performed independently for each candidate ANN. Therefore, the parallel performance of this step scales with hardware capabilities when there are enough candidates to occupy the entire device.

% ----------------------------------------------------------------------------
\subsection{Computation of Number of Truly Positive Top 20\% Instances} \label{implementation}
% ----------------------------------------------------------------------------
To compute the number of truly positive instances in the top 20\% of scores, if training instances are unsorted, we must read from memory each instance's true class and corresponding ANN output value and compare the output value to the candidate's threshold. For $n$ training instances, this requires $2 * n$ memory accesses. However, by keeping the $n_p$ positive instances first in memory, we need only read the first $n_p$ ANN outputs, knowing that they correspond to the positive instances. This reduces the number of memory accesses to $n_p$, which is significant particularly for training data with low positive rate (in our case, $2 * n = 189364$ and $n_p = 2080$). Sorting the training instances by class is a cheap operation and can be done by the host CPU as part of initialization.

Counting of positive instances above the candidate's threshold is performed independently, in parallel, for each candidate. However, because there may not be enough individuals to fully occupy the GPU, we also split the counting process for each individual into 256 independent sub-counts: the first sub-counter processes training instances 0, 256, 512, \ldots, the second sub-counter instances 1, 257, 513, \ldots, such that the sub-counts can be executed in parallel. The striped assignment of training instance indices to sub-counters is necessary to take advantage of the full memory bandwidth of the GPU. The host CPU then aggregates the sub-counts by summing, which is a very lightweight operation.

Finally, each count is divided by the number of positive instances to give a positive rate in the top 20\%. Then, the ratio of this positive rate to the overall positive rate is the lift of each individual. These operations are also comparatively lightweight and executed on the host CPU.

@@

\begin{table}[h]
	\centering
	\begin{tabular}{lllll}
	\textbf{Legend Label} & \textbf{Hardware} & \textbf{Clock} & \textbf{Free RAM} & \textbf{OS} \\
	\hline
	Core2  & Intel Core 2 Q9450 CPU & 2.66~GHz & 7.5~GB & Linux 2.6 (64-bit) \\
	i7     & Intel i7 920 CPU       & ?        & ?      & Windows 7 (64-bit) \\
	GTX260 & NVidia GTX260 GPU      & ?        & N/A    & Windows 7 (64-bit) \\
	GTX275 & NVidia GTX275 GPU      & 1.4~GHz  & N/A    & Linux 2.6 (64-bit) \\
	\hline
	\end{tabular}
	\caption{Experimental hardware. Free RAM is OS-reported available memory before running each experiment.}
	\label{tab:experimental-hardware}
\end{table}

% ----------------------------------------------------------------------------
\subsection{Performance Results} \label{performance}
% ----------------------------------------------------------------------------
In most cases, the GPU implementation resulted in approximately an order of magnitude speedup over a CPU implementation (Figure~\ref{fig:training-performance}). For population sizes below 10, observed speedup decreases and no speedup is observed in the single-individual case. This is due to GPU initialization and data copying dominating execution time. However, in these cases, both implementation are still quite fast (20~ms per generation).

\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth]{fig-performance}
	\caption{Training performance as a function of population size (left) and training set size (right). Note logarithmic axes.}
	\label{fig:training-performance}
\end{figure}

% ############################################################################
\section{Conclusion} \label{concl}
% ############################################################################
We have presented an implementation of GA training of ANN classifiers for the CUDA platform for GPU programming. By carefully designing memory organization, algorithm computational load and memory access patterns, we have obtained a 10-fold speedup compared to a conventional sequential CPU implementation. A multipass approach was required for part of our system to compensate for the limited amount of fast on-chip GPU memory. Our method scales across population and training set sizes and is expected to be useful in other data intensive machine learning and data mining tasks.
